Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex
Edits with Hugues.

Our $\globalbarrier$ primitive is a kernel-level function: if it appears in
conditional code then it must be reached by \emph{all} threads
executing the cooperative kernel.  On reaching a $\globalbarrier()$, a
thread waits until all threads have arrived at the barrier.  Once all
threads have arrived, the threads may proceed past the barrier with
the guarantee that all global memory accesses issued before the
barrier have completed.

The $\resizingglobalbarrier$ primitive is also a kernel-level
function.  It is identical to $\globalbarrier$, except that it caters
for cooperation with the scheduler: by issuing a
$\resizingglobalbarrier$ the programmer indicates that the cooperative
kernel is prepared to proceed after the barrier with fewer workgroups,
or with additional workgroups joining the computation.
Semantically, it is as if every workgroup executes $\offerkill$ on reaching the
barrier, and every workgroup executes $\offerfork$ on leaving the
barrier.  Thus $\resizingglobalbarrier()$ is semantically equivalent
to: \ADComment{Does the order of these three calls matter, semantically?}
%
\lstset{basicstyle=\tt}
\begin{lstlisting}
  $\offerkill()$;
  $\globalbarrier()$;
  $\offerfork()$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

The $\globalbarrier$ primitive can be implemented following any
inter-workgroup barrier design, e.g.~\cite{...}, and the atomic
operations provided by the OpenCL 2.0 memory model enable a
memory-safe implementation~\cite{...}.  However, implementing such a
barrier is involved, hence why we include this function as a
primitive.  As noted above, the $\resizingglobalbarrier$ primitive
is, semantically, syntactic sugar for a sequence of other primitive
calls.  We include it as a primitive because (a) we have found that
global barriers often provide a natural point for cooperation with the
scheduler, and (b) a custom implementation of
$\resizingglobalbarrier()$ that provides the semantics described above
can be made significantly more efficient than the above sequence of
calls.  We discuss our efficient implementation of
$\resizingglobalbarrier$ in detail in
Section~\ref{sec:resizingbarrier}.

@@

An OpenCL programmer writing a cooperative kernel should pay careful
attention to the following details.

\paragraph{A changing number of workgroups}

In regular OpenCL, $\getnumgroups{0}$ returns a runtime-constant
value: once the number of executing workgroups has been identified, it
is guaranteed not to change.

In contrast, the number of workgroups executing a cooperative kernel
can vary due to workgroups executing $\offerkill$, $\offerfork$ and
$\resizingglobalbarrier$.  The programmer must therefore write their
cooperative kernel in a manner that is robust to changes in the value
returned by $\getnumgroups{0}$.

@@

\ADComment{Need to account for the issue that $M$ might decrease as
  workgroups actually exit the kernel.}

If a cooperative kernel does not directly issue $\offerkill$ and
$\offerfork$ calls then the only possibility for the number of
workgroups to change is due to $\resizingglobalbarrier$ being invoked.

@@

Kernel state: an $N$-tuple $(\sigma, (w_1, w_2, \dots, w_M, \bot,
\dots, \bot))$, where each $w_i$ is a workgroup state, which are
followed by $N-M$ occurrences of $\bot$ to indicate absent workgroups.

We treat thread-level semantics abstractly by assuming a transition
relation $(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l',
\mathit{ss}'))$ that is defined unless $\mathit{ss} =
\mathsf{special}(); \mathit{tt}$, where $\mathit{special}$ is one of
$\offerkill$, $\offerfork$, $\globalbarrier$ or
$\resizingglobalbarrier$.

To abstractly account for the interaction between global barrier
synchronization and the GPU memory model, we assume a $\mathsf{sync}$
that can be applied to the shared state $\sigma$, or to the shared
state and a thead local state.  In the first case,
$\mathsf{sync}(\sigma)$ returns an updated shared state that reflects
all in-flight memory stores having completed.  In the second case,
$\mathsf{sync}(\sigma, l)$ returns a thread local state reflecting all
in-flight memory loads issued by the thread having completed.

@@

\[
\inferrule{
\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
\\
w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
}
{
(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
}
(\textsc{Fork-No-Op})
\]

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\mathit{ss})
\\
\forall i \;.\;\forall j\;.\; l_{i,j}' = \mathsf{sync}(\sigma, l_{i,j})
\\
\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}', \mathit{ss})
\\
\sigma' = \mathsf{sync}(\sigma)
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma', (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Barrier})
\]

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
\\
\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \offerkill(); \globalbarrier(); \offerfork();\mathit{ss})
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Resizing-Barrier})
\]

>>>>

Our $\globalbarrier$ primitive is a kernel-level function: if it
appears in conditional code then it must be reached by \emph{all}
threads executing the cooperative kernel.  On reaching a
$\globalbarrier()$, a thread waits until all threads have arrived at
the barrier.  Once all threads have arrived, the threads may proceed
past the barrier with the guarantee that all global memory accesses
issued before the barrier have completed.  The $\globalbarrier$
primitive can be implemented following any inter-workgroup barrier
design, e.g.~\cite{...}, and the atomic operations provided by the
OpenCL 2.0 memory model enable a memory-safe
implementation~\cite{...}.  However, implementing such a barrier is
involved, hence why we include this function as a primitive.

The $\resizingglobalbarrier$ primitive is also a kernel-level
function.  It is identical to $\globalbarrier$, except that it caters
for cooperation with the scheduler: by issuing a
$\resizingglobalbarrier$ the programmer indicates that the cooperative
kernel is prepared to proceed after the barrier with fewer workgroups,
or with additional workgroups joining the computation.

Semantically, when all threads have reached $\resizingglobalbarrier$,
the number of active workgroups, $M$, is set to a new value, $M'$ say.
If $M' = M$ then the active workgroups remain unchanged and proceed to
execute after the barrier.  If $M' < M$, workgroups $[M', M-1]$ are
killed.  If $M' > M$ then $M'-M$ new workgroups join the computation,
as if they were forked from workgroup 0.  In particular, the
$\transmit$-annotated local state of thread 0 in workgroup 0 is
transmitted to the threads of the new workgroups.  The change in
number of active workgroups from $M$ to $M'$ is atomic.

The semantics of $\resizingglobalbarrier$ can be modelled via the following sequence of calls:

\lstset{basicstyle=\tt}
\begin{lstlisting}
  $\globalbarrier()$;
  if($\getgroupid{0}$ != 0) $\offerfork$();
  $\globalbarrier()$;
  $\offerkill()$;
  $\globalbarrier()$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

The enclosing $\globalbarrier$ calls make the operation appear to be
atomic.  The middle $\globalbarrier$ is required to ensure that if
$M'\geq M$, the workgroups with ids smaller than $M$ are left intact.

Because $\resizingglobalbarrier$ can be implemented in the above
manner, we do not regard it conceptually as a primitive of the
cooperative kernels programming model.  However, we discuss in
Section~\ref{sec:resizingbarrier} how $\resizingglobalbarrier$ can be
implemented significantly more efficiently if it is treated as a
primitive operation that interacts directly with the scheduler.

@@

We remark on some issues that programmer working with cooperative
kernels should be aware of.

\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel.  Although $\getnumgroups()$ has
atomic semantics, and so can be called at any time to yield a
well-defined value, the value returned by two successive calls may
differ if workgroups have executed $\offerkill$, $\offerfork$ or
$\resizingglobalbarrier$ in-between.  The programmer must therefore
write their cooperative kernel in a manner that is robust to changes
in the value returned by $\getnumgroups()$.

@@

The situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.

 then the only possibility for the number of
workgroups to change is due to $\resizingglobalbarrier$ being invoked.

@@

Kernel state: a pair $(\sigma, (w_1, w_2, \dots, w_M, \bot, \dots, \bot))$, where $\sigma$ is the shared state,
each $w_i$ is a workgroup state, which are
followed by $N-M$ occurrences of $\bot$ to indicate absent workgroups.

We treat thread-level semantics abstractly by assuming a transition
relation $(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l',
\mathit{ss}'))$ that is defined unless $\mathit{ss} =
\mathsf{special}(); \mathit{tt}$, where $\mathsf{special}$ is one of
$\offerkill$, $\offerfork$, $\globalbarrier$ or
$\resizingglobalbarrier$.

To abstractly account for the interaction between global barrier
synchronization and the GPU memory model, we assume a function
$\mathsf{sync}$ that is applied to a global state.  This returns a new
global state in which the shared state and thread local states have
been updated to reflect the fact that all stores to memory and loads
from memory to thread-local storate have completed.

@@

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\mathit{ss})
\\
\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \mathit{ss})
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow \mathsf{sync}(\sigma', (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Barrier})
\]

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
\\
\forall i \neq 0 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
\\
\forall j\;.\;w_0'(j) = (l_{0,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Resizing-Barrier})
\]
